Avoiding tipping points in the management of ecological systems: a non-parametric Bayesian approach

### 2013-06-26 inst/doc/nonparametric-bayes.tex
updated draft

Decision making under uncertainty is a ubiquitous challenge of natural
resource management and conservation.

Here we illustrate how a stochastic dynamic programming algorithm can be
driven by the predictions from a Gaussian process model, sidestepping
the need for an accurate model-based description of the sustem dynamics.

@@

Traditional approaches to optimal control (Pontryagin's principle,
stochastic dynamic programming) rely on knowledge of the state equation.
Such simple parametric models, frequently justified by mechanistic
underpinnings are used to provide probabilities of future states.
Fisheries literature, which enjoys some of the longest history in the
use of the optimal control framework, provides a clear example. Models
such as the Ricker or Beverton Holt curves are ubiquitous {[}too strong?
references?{]} Parametric models can reflect variability introduced by a
stochastic environment or demographic process my incorporating random
shocks from a given distribution.

@@

@@

@@

We demonstrate how a Gaussian Process model of stock recruitment can
lead to nearly optimal management through stochastic dynamic
programming, comperable to knowing the correct structural equation for
the underlying simulation. Meanwhile, parametric models that do not
match the underlying dynamics can perform very poorly, even though they
fit the data as well as the true model.

Ecological research and management strategy should pay closer attention
to the opportunities and challenges nonparametric modeling can offer.

@@

\subsubsection{Terminology: ``Non-parametric''}

@@

@@

Both parametric and nonparametric approaches will require training data
on which to base their model of the process. We generate the training
data under the model described in Eq 1 for $T_{\textrm{obs}}=40$ time
steps, under a known but not necessarily optimal sequence of harvest
intensities, $h_t$. For simplicity we imagine a fishery that started
from zero harvest pressure and has been gradually increasing the
harvest.

\textbf{\emph{Should we include any emprical examples?}}

\subsubsection{The optimal control problem in fisheries management}

@@

We generate the data with a four-parameter model that contains a tipping
point (an Allee effect, see {[}@Allen, @Courchamp{]}) below which the
stock decreases to zero,

@@

@@

The essence of the GP approach can be captured in the following thought
experiment: An exhaustive parametric approach to the challenge of
structural uncertainty might proceed by writing down all possible
functional forms for the underlying dynamical system with all possible
parameter values for each form, and then consider searching over this
huge space to select the most likely model and parameters; or using a
Bayesian approach, assign priors to each of these possible models and
infer the posterior distribution of possible models. The GP approach can
be thought of as a computationally efficient approximation to this
approach. GPs represent a large class of models that can be though of as
capturing or reasonably approximating the set of models in this
collection. By modeling at the level of the process, rather than the
level of parametric equation, we can more concisely capture the possible
behavior of these curves. In place of a parametric model of the
dynamical system, the GP approach postulates a prior distribution of
(n-dimensional) curves that can be though of as approximations to a
range of possible (parametric) models that might describe the data. The
GP allows us to consider a set of possible curves simultaneously.

\subsubsection{Background on Gaussian Process inference}

@@

This completes the algorithm adapting the GP to the sequential
decision-making problem through SDP, which we believe has not yet been
demonstrated in the literature. We provide an R package implementation
of this, along with the Gaussian process inference, in the supplemental
materials.

@@

The posterior distribution for the hyper-parameters of the Gaussian
process model are estimated by Metropolis-Hastings algorithm, again with
details and code provided in the Appendix. Rasmussen and Williams (2006)
provides an excellent general introduction to Gaussian Processes and
Munch et al. (2005) first discusses their application in this context.

@@

@@

@@

@@

 Data complexity. Perhaps too far out of scope\ldots{} The nonparametric
Bayesian approach is also better suited to complex and disparate data.
Incorporating various sources of information into mechanistic models can
be an immensely difficult due to the increased complexity involved.

@@

@@

@@

@@

>>>>

\paragraph{Opening}

Decision making under uncertainty is a ubiquitous challenge of natural
resource management and conservation.\\Here we illustrate how a
stochastic dynamic programming algorithm can be driven by the
predictions from a Gaussian process model, sidestepping the need for an
accurate model-based description of the system dynamics.

add references to {[}@Fischer2009, @Polasky2011{]} as persective on
combining complex / critical dynamics + control theory

\paragraph{Models for decision-making under uncertainty}

@@

More historical context, e.g. (Reed 1979, )

\paragraph{Parametric models}

Traditional approaches to optimal control (Pontryagin's principle,
stochastic dynamic programming) rely on knowledge of the state equation.
Such simple parametric models, frequently justified by mechanistic
underpinnings are used to provide probabilities of future states.
Fisheries literature, which enjoys some of the longest history in the
use of the optimal control framework, provides a clear example {[} {]} .
Models such as the Ricker or Beverton Holt curves are ubiquitous {[}too
strong? references?{]} Parametric models can reflect variability
introduced by a stochastic environment or demographic process my
incorporating random shocks from a given distribution.

\paragraph{Parametric uncertainty}

@@

(Mangel 1985; Schapaugh and Tyre 2013)

\paragraph{Structural Uncertainty}

@@

(B. K. Williams 2001; Athanassoglou and Xepapadeas 2012).

\paragraph{Comparing models}

@@

We demonstrate how a Gaussian Process model of stock recruitment can
lead to nearly optimal management through stochastic dynamic
programming, comperable to knowing the correct structural equation for
the underlying simulation. Meanwhile, parametric models that do not
match the underlying dynamics can perform very poorly, even though they
fit the data as well as the true model.\\Ecological research and
management strategy should pay closer attention to the opportunities and
challenges nonparametric modeling can offer.

\paragraph{Background on Fisheries Context}

\paragraph{Background on Tipping points}

@@

@@

\subsection{The optimal control problem in fisheries management}

@@

\paragraph{Simulated training data}

We generate initial observational data under the model described in Eq 1
for $T_{\textrm{obs}}=40$ time steps, under a given arbitrary sequence
of harvest intensities, $h_t$. We consider the case in which most of the
data comes from a limited region of state space (e.g.~near a stable
equilibrium), leaving us without observations of the population dynamics
at very low levels which would be useful in discrimating between
recruitment curves {[}@{]} or demonstrating the existence of a tipping
point (Scheffer et al. 2001).\\Using data simulated from a specified
model rather than empirical data permits the comparison against the true
underlying dynamics, setting a bar for the optimal performance possible.

@@

We generate the data with a four-parameter model that contains a tipping
point, as discussed above (equation 1), (an Allee effect, see {[}@Allen,
@Courchamp{]}) below which the stock decreases to zero,

@@

\paragraph{Terminology: ``Non-parametric''}

@@

The essence of the GP approach can be captured in the following thought
experiment: An exhaustive parametric approach to the challenge of
structural uncertainty might proceed by writing down all possible
functional forms for the underlying dynamical system with all possible
parameter values for each form, and then consider searching over this
huge space to select the most likely model and parameters; or using a
Bayesian approach, assign priors to each of these possible models and
infer the posterior distribution of possible models. The GP approach can
be thought of as a computationally efficient approximation to this
approach. GPs represent a large class of models that can be though of as
capturing or reasonably approximating the set of models in this
collection. By modeling at the level of the process, rather than the
level of parametric equation, we can more concisely capture the possible
behavior of these curves. In place of a parametric model of the
dynamical system, the GP approach postulates a prior distribution of
(n-dimensional) curves that can be though of as approximations to a
range of possible (parametric) models that might describe the data. The
GP allows us to consider probabilities on a large set of possible curves
simultaneously.

The posterior distribution for the hyper-parameters of the Gaussian
process model are estimated by Metropolis-Hastings algorithm, again with
details and code provided in the Appendix. Rasmussen and Williams (2006)
provides an excellent general introduction to Gaussian Processes and
Munch et al. (2005) first discusses their application in the context of
population dynamics models such as fisheries stock-recruitment
relationships.

\subsubsection{SDP via GP}

@@

This completes the algorithm adapting the GP to the sequential
decision-making problem through SDP, which has not been previously
demonstrated.\\We further provide an R package implementation as
described in the supplemental materials.

@@

@@

\paragraph{Big picture: Linking GP to SDP}

@@

Though simple mechanistically motivated models may be best for the
latter case (@Cuddington2013; Geritz and Kisdi 2012), such models can be
not only inaccurate but misleading in quantitative decision making. This
arises for two reasons:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\itemsep1pt\parskip0pt\parsep0pt
\item
  We do not know what the correct parameter values are for the models.
\item
  We do not know what the correct models are for ecological systems.
\end{enumerate}

The complexity of ecological interactions and a lack of data contribute
greatly to both of the problems. This concern is particularly acute in
case of ecological tipping points (Barnosky et al. 2012; Scheffer et al.
2001), which arise from feedbacks common in ecological systems and can
lead to a sudden catastrophic transition to an undesirable state. In
general we do not know where such tipping points are unless we have
already observed the transition, in which case it is frequently too late
to respond. Though there may exist early warning signals for certain
kinds of these transitions that are driven by slow changes (Scheffer et
al. 2009), we do not know when, where, or how to apply them to the
decision making context more generally (Boettiger and Hastings 2013).

@@

\subsubsection{Higher dimensions}

@@

@@

Barnosky, Anthony D., Elizabeth a Hadly, Jordi Bascompte, Eric L.
Berlow, James H. Brown, Mikael Fortelius, Wayne M. Getz, et al. 2012.
``Approaching a state shift in Earth's biosphere.'' \emph{Nature} 486
(7401) (jun): 52--58. doi:10.1038/nature11018.
\url{http://www.nature.com/doifinder/10.1038/nature11018}.

@@

Boettiger, Carl, and Alan Hastings. 2013. ``Tipping points: From
patterns to predictions.'' \emph{Nature} 493 (7431) (jan): 157--158.
doi:10.1038/493157a.
\url{http://www.nature.com/doifinder/10.1038/493157a}.

@@

Geritz, Stefan A. H., and Eva Kisdi. 2012. ``Mathematical ecology: why
mechanistic models?.'' \emph{Journal of mathematical biology} 65 (6-7)
(dec): 1411--5. doi:10.1007/s00285-011-0496-3.
\url{http://www.ncbi.nlm.nih.gov/pubmed/22159789}.

@@

Scheffer, Marten, Jordi Bascompte, William A. Brock, Victor Brovkin,
Stephen R. Carpenter, Vasilis Dakos, Hermann Held, Egbert H. van Nes,
Max Rietkerk, and George Sugihara. 2009. ``Early-warning signals for
critical transitions.'' \emph{Nature} 461 (7260): 53--9.
\url{http://www.ncbi.nlm.nih.gov/pubmed/19727193}.
